---
title: "Methodologies mean_GRM, SD_GRM and GRM_SVM"
author: "Hélène Wilmot et al"
date: "27 April 2023"
output: word_document
---

The objective of this R script is to explain the development of the different methodologies developed. 
These methodologies are based on the genomic relationship matrix (GRM).
Mean_GRM refers to the assignment based on the highest mean relatedness of an animal to the breeds found in the reference set.
SD_GRM refers to the assignment based on the highest standard deviation (SD) of the relatedness of an animal to the breeds found in the reference set.
GRM_SVM refers to the assignment based on a linear support vector machine (SVM) using the different values of mean and SD of the relatedness of the animal to be assigned as an input.
For the PLS_NSC methdology, please refer to the Supplemental material of Wilmot et al. (2022): https://doi.org/10.1111/jbg.12643

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Working directory should be set

# Used packages

```{r}
#install.packages("dplyr",repos = "http://cran.us.r-project.org")
library(dplyr)
#install.packages("caret",repos = "http://cran.us.r-project.org")
library(caret) #To cross-validate and validate GRM_SVM
#install.packages("matrixStats",repos = "http://cran.us.r-project.org")
library(matrixStats)
```

# 1. Mean_GRM

## 1.1. Computation of mean relatedness of each validation animal to each breed of the reference set

```{r}
#GRM computed with calc_grm software
GRM_valid<-read.table(file="G.grm", header=FALSE)
head(GRM_valid)
#V1 and V2 are number order assigned to animals , V3 is the relatedness, V4 and V5 are number IDs (here identical to V1 and V2)
GRM_valid<-subset(GRM_valid, V1>325 & V2<326) #Keep relatedness of validation animals to reference animals only

#Get the IDs of the reference animals
ID_RS1<-read.table(file=paste("1_RS1_ID.txt", sep=""), header=FALSE, sep="") #V1 is the breed, V2 is the Interbull ID and V3 is the number ID  
ID_EBRW_RS1<-subset(ID_RS1, V1=="EBRW") #IDs of EBRW reference animals
ID_MRY_RS1<-subset(ID_RS1, V1=="MRY") #IDs of MRY reference animals
ID_RPO_RS1<-subset(ID_RS1, V1=="RPO") #IDs of RPO reference animals

GRM_valid_EBRW<-dplyr::filter(GRM_valid, V2 %in% ID_EBRW_RS1$V3) #Keep relatedness of validation animals to the EBRW reference set
mean_to_EBRW<- GRM_valid_EBRW %>% group_by(V1) %>% 
    dplyr::summarise(mean_relatedness_EBRW=mean(V3)) #Compute the mean relatedness of validation animals to the EBRW reference set
head(mean_to_EBRW)

GRM_valid_MRY<-dplyr::filter(GRM_valid, V2 %in% ID_MRY_RS1$V3) #Keep relatedness of validation animals to the MRY reference set
mean_to_MRY<- GRM_valid_MRY %>% group_by(V1) %>% 
    dplyr::summarise(mean_relatedness_MRY=mean(V3)) #Compute the mean relatedness of validation animals to the MRY reference set

GRM_valid_RPO<- dplyr::filter(GRM_valid, V2 %in% ID_RPO_RS1$V3) #Keep relatedness of validation animals to the RPO reference set
mean_to_RPO<- GRM_valid_RPO %>% group_by(V1) %>% 
    dplyr::summarise(mean_relatedness_RPO=mean(V3)) #Compute the mean relatedness of validation animals to the RPO reference set
``` 

## 1.2. Assignment of validation animals to their breed based on mean_GRM

```{r}
#Merging mean relatedness of validation animals to each breed of interest
mergedrel<-merge(mean_to_EBRW, mean_to_MRY, by=intersect("V1","V1"))
mergedrel2<-merge(mergedrel, mean_to_RPO, by=intersect("V1","V1"))
#Giving the real breed of origin of validation animals
mergedrel2$Breed<-rep(c("EBRW","RPO","MRY"),c(113,66,146))
mergedrel2 <- transform(mergedrel2, predicted=c("EBRW", "MRY", "RPO")[max.col(mergedrel2[-c(1,5)])]) #Predict the breed of origin of validation animals based on the highest mean relatedness
#Determine if predicted breed is the same as the breed of origin
mergedrel2$diff<-ifelse(mergedrel2$Breed == mergedrel2$predicted, 1,0)
head(mergedrel2)
percentage<-(sum(mergedrel2$diff)/nrow(mergedrel2)) #Percentage of correct assignment=global accuracy
percentage
``` 

# 2. SD_GRM
## 2.1. Computation of SD of the relatedness of each validation animal to each breed of the reference set

```{r}
#If performed alone, same steps as mean_GRM must be computed before computation of the SD of the relatedness of validation animals
sd_to_EBRW<- GRM_valid_EBRW %>% group_by(V1) %>% 
dplyr::summarise(sd_relatedness_EBRW=sd(V3)) #Compute the SD of the relatedness of validation animals to the EBRW reference set
head(sd_to_EBRW)
sd_to_MRY<- GRM_valid_MRY %>% group_by(V1) %>% 
dplyr::summarise(sd_relatedness_MRY=sd(V3)) #Compute the SD of the relatedness of validation animals to the MRY reference set
sd_to_RPO<- GRM_valid_RPO %>% group_by(V1) %>% 
dplyr::summarise(sd_relatedness_RPO=sd(V3)) #Compute the SD of the relatedness of validation animals to the RPO reference set
```
  
## 1.2. Assignment of validation animals to their breed based on SD_GRM
```{r}
#Merging SD of the relatedness of validation animals to each breed of interest
mergedrelSD<-merge(sd_to_EBRW, sd_to_MRY, by=intersect("V1","V1"))
mergedrelSD2<-merge(mergedrelSD, sd_to_RPO, by=intersect("V1","V1"))
#Giving the real breed of origin of validation animals
mergedrelSD2$Breed<-rep(c("EBRW","RPO","MRY"),c(113,66,146))
mergedrelSD2 <- transform(mergedrelSD2, predicted=c("EBRW", "MRY", "RPO")[max.col(mergedrelSD2[-c(1,5)])])#Predict the breed of origin of validation animals based on the highest SD of the relatedness
#Determine if predicted breed is the same as the breed of origin
mergedrelSD2$diff<-ifelse(mergedrelSD2$Breed == mergedrelSD2$predicted, 1,0)
head(mergedrelSD2)
percentageSD<-(sum(mergedrelSD2$diff)/nrow(mergedrelSD2))#Percentage of correct assignment=global accuracy
percentageSD
```

# 3. GRM_SVM
## 3.1. Formating
Transform the table obtained in calc_grm as a matrix, easier to handle for the GRM_SVM methodology

```{r}
IN <- data.table::fread("G.grm") #Read the GRM table
# Column in V1
# Row in V2
n.entries <- max(IN$V5) #To determine the dimension of the GRM matrix
m <- matrix(9, nrow = n.entries, ncol = n.entries) #Set an empty matrix with the correct dimensions
#A loop to fill the matrix with the values of the GRM table
counter <- 1
for(col.i in 1:ncol(m)){
for(row.j in 1:col.i){
m[row.j, col.i] <- IN$V3[counter]
m[col.i, row.j] <- IN$V3[counter]
counter <- counter + 1
}
}
m<-m[1:325,1:325] #Keep reference animals
colnames(m)<-c(1:325) #Assign the ID of animals as column names
diag(m)<-NA #Replace self-relatedness (on the diagonal) by NA to avoid a bias in the computation of mean relatedness
m[1:5,1:5]
```  

## 3.2. Computation of mean and SD of the relatedness within the reference set (self-relatedness excluded)

```{r}
EBRW_list<-ID_EBRW_RS1$V3 #List of EBRW reference animals
MRY_list<-ID_MRY_RS1$V3 #List of MRY reference animals
RPO_list<-ID_RPO_RS1$V3 #List of RPO reference animals
#Mean relatedness to EBRW animals within RS1
forComSVMEBRW<-as.data.frame(rowMeans(subset(m,select=EBRW_list),na.rm=TRUE)) 
forComSVMEBRW$ID<-rownames(forComSVMEBRW)#Set IDs in the file
head(forComSVMEBRW)

#Mean relatedness to MRY animals within RS1
forComSVMMRY<-as.data.frame(rowMeans(subset(m,select=MRY_list),na.rm=TRUE)) 
forComSVMMRY$ID<-rownames(forComSVMMRY) #Set IDs in the file

#Mean relatedness to RPO animals within RS1
forComSVMRPO<-as.data.frame(rowMeans(subset(m,select=RPO_list),na.rm=TRUE))
forComSVMRPO$ID<-rownames(forComSVMRPO)#Set IDs in the file

#SD of the relatedness to EBRW animals within RS1
forComSVMEBRWsd<-as.data.frame(rowSds(subset(m,select=EBRW_list),na.rm=TRUE))
forComSVMEBRWsd$ID<-rownames(forComSVMEBRWsd)#Set IDs in the file
head(forComSVMEBRWsd)

#SD of the relatedness to MRY animals within RS1
forComSVMMRYsd<-as.data.frame(rowSds(subset(m,select=MRY_list),na.rm=TRUE))
forComSVMMRYsd$ID<-rownames(forComSVMMRYsd)#Set IDs in the file

#SD of the relatedness to RPO animals within RS1
forComSVMRPOsd<-as.data.frame(rowSds(subset(m,select=RPO_list),na.rm=TRUE))
forComSVMRPOsd$ID<-rownames(forComSVMRPOsd)#Set IDs in the file

#Merge the six variables (mean and SD of the relatedness within RS1)
combi1 <- merge(forComSVMEBRW, forComSVMMRY, by=intersect("ID","ID"))
combi2<-merge(forComSVMRPO,combi1, by=intersect("ID","ID"))
combi3<-merge(forComSVMEBRWsd,combi2, by=intersect("ID","ID"))
combi4<-merge(forComSVMMRYsd,combi3, by=intersect("ID","ID"))
combi5<-merge(forComSVMRPOsd,combi4, by=intersect("ID","ID"))
#Define column names
colnames(combi5)<-c("ID","sd_RPO", "sd_MRY","sd_EBRW","relatedness_RPO","relatedness_EBRW","relatedness_MRY") 
combi5$ID<-as.numeric(combi5$ID) #Define ID as numeric
combi5<-combi5[order(combi5$ID),] #Order by ID
combi5$Breed<-ID_RS1$V1 #Define breed of origin
combi_final<-combi5[,c(1,6,7,5,4,3,2,8)] #Order columns (important to have the same order for validation)
combi_final$Breed<-as.factor(combi_final$Breed) #Define breed as a factor
levels(combi_final$Breed) #Check the factor levels of the breed
head(combi_final)
```
 
## 3.3. Training of the linear SVM on the values of the reference set

```{r}
set.seed(1001) # To fix the algorithm
ctrl<-trainControl(method="repeatedcv", number=10, classProbs=TRUE, selectionFunction= "best") # Parameter optimized based on the best value obtained in 10 fold cross-validation

grid<-expand.grid(C=c(0.001,0.01,0.1,0.2,0.3,0.4,0.5, 0.6, 0.7, 0.8, 0.9)) #Cost values to be tested
row.names(combi_final)<-combi_final$ID #Set ID in rownames
combi_final_svm<-combi_final[,-1] #Remove the variable ID
svmL = train(Breed~.,data=combi_final_svm,
               method="svmLinear", 
               trControl = ctrl,
               preProc=c("center","scale"), #Mean centring and SD scaling for each of the six variables
               metric="Accuracy",
               tuneGrid=grid)
svmL
```  
  
## 3.4. Validation of the linear SVM on the validation set

```{r}
#Merge files with the mean and SD of the relatedness of the validation set
#These were previously computed on sections 1.1. and 2.1.
combination<-merge(mergedrel2, mergedrelSD2, by=intersect("V1","V1"))
combination<-combination[,c(1:4,8:11)] #Same order of columns as for the input file used for training
colnames(combination)<-c("ID","relatedness_EBRW","relatedness_MRY","relatedness_RPO","sd_EBRW","sd_MRY","sd_RPO","Breed") #Define column names as previously
combination$Breed<-as.factor(combination$Breed) #Define breed as a factor
combination_svm<-combination[,-1] #Remove the variable ID
levels(combination_svm$Breed) #Check the factor levels of the breed
head(combination_svm)
pred<-predict(svmL, combination_svm) #Predict the breed based on the model tuned by cross-validation (section 3.3.)
confus<-confusionMatrix(pred, combination_svm[,7])
confus #Confusion matrix with global accuracy, sensitivities and specificities for each breed
```  
